The evolution of gene expression in the term placenta of viviparous mammals

### 2016-01-12 manuscript/manuscript.Rnw
discuss PRL in discussion

The elephant is strikingly the only mammal analyzed that shows high
expression of the primarily pituitary hormone prolactin (PRL) in the
placenta. %
\DLA{This is totally not true. \textit{A. fusciceps} expresses PRL,
  and so does \textit{S. carmeli}, and \textit{B. taurus}, \textit{M.
    musculus}, all have expansions of prolatacin.}
%
All afrotherians yet analyzed contain only one prolactin
gene, and prolactin has undergone adaptive evolution on the elephant
lineage \cite{Wallis2009:ProlactininAfrotheriacharacterization}.
\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
presence of a prolactin in the trophoblast of the elephant placenta
using immunohistological staining and human derived antibodies. A
previous analysis of the elephant placenta also detected the
expression of prolactin (Hou et al., 2012). Here we confirm that the
sole prolactin in the elephant is not only expressed but highly
expressed in the placenta, and likely functions both in the
maintenance of the corpus luteum and as a placental lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}. PRL is not expressed in the opossum, even though there
is evidence that it is expressed in the placenta of another marsupial,
the tammar wallaby \cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests that
placental prolactin expression in marsupials may be as diverse as in
eutherian mammals. Although, PRL is not expressed to any great degree
in the placenta other mammals, we see very high expression of genes
which resulted from duplications of PRL in both the cow \DLA{(Table X)}, as
well as mouse \DLA{(Supplemental Table X)}. \DLA{Describe these}

The human gene CGA codes for the alpha subunit of multiple hormones
including pituitary hormones (luteinizing hormone,
The \ac{GH} family has expanded in anthropoid primates, resulting in
approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
reliably distinguish transcripts between the two). All but
\textit{GH1} show substantial expression in the \human placenta
(\cref{fig:csh1_evolution_tree}). There is some evidence that
increased expression of these genes may be correlated with overall
birth weight of offspring in \human
\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
possible that the extreme expression levels of these genes may be due
to high birth weight. \human placental \ac{GH} genes are involved in
increasing blood flow, replacing maternal pituitary expression of
\ac{GH}, as well as trophoblast invasion
\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
and it is likely that they have the same function in \bonobo.

>>>>

The elephant is strikingly the only mammal analyzed that shows high
expression of the primarily pituitary hormone prolactin (PRL) in the
placenta. All afrotherians yet analyzed contain only one prolactin
gene, and prolactin has undergone adaptive evolution on the elephant
lineage \cite{Wallis2009:ProlactininAfrotheriacharacterization}.
\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
presence of a prolactin in the trophoblast of the elephant placenta
using immunohistological staining and human derived antibodies. A
previous analysis of the elephant placenta also detected the
expression of prolactin
\cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}. Here we
confirm that the sole prolactin in the elephant is not only expressed
but highly expressed in the placenta, and likely functions both in the
maintenance of the corpus luteum and as a placental lactogen
\cite{Yamamoto.ea2011:PlacentationinAfricanelephant}. \textit{PRL} is
not expressed in the opossum, even though there is evidence that it is
expressed in the placenta of another marsupial, the tammar wallaby
\cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
that placental prolactin expression in marsupials may be as diverse as
in eutherian mammals. Although, \textit{PRL} is not expressed to any
great degree in the placenta other mammals, we see very high
expression of genes which resulted from duplications of \textit{PRL}
in the \mouse, \galili, \carmeli, and \cow
(\cref{fig:prolactin_evolution_tree}).

The \ac{GH} family has expanded in anthropoid primates, resulting in
approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
reliably distinguish transcripts between the two). All but
\textit{GH1} show substantial expression in the \human placenta
(\cref{fig:csh1_evolution_tree}). There is some evidence that
increased expression of these genes may be correlated with overall
birth weight of offspring in \human
\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
possible that the extreme expression levels of these genes may be due
to high birth weight. \human placental \ac{GH} genes are involved in
increasing blood flow, replacing maternal pituitary expression of
\ac{GH}, as well as trophoblast invasion
\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
and it is likely that they have the same function in \bonobo.
